Thesis: draft

### 2021-03-25 draft.tex
wrote part on network models in chapter 5, NOT FINISHED, still need to include shortcommings of the models

\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bounded by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downfalls in the sense that people might get trapped in "echo chambers" that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate which network properties (eg. community structure, clustering coefficient, etc) may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. Beside that, we will also pay some attention to the possible ways of distributing the opinions; we will compare randomly distributed cases to cases where entire communities are given one single opinion.
% definetely not finished!! --> add some type of conclusions or not? talk about running model on real world networks?
\end{abstract}

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.
% refer to relevant sections in introduction?

\section{Complex networks}

@@

\section{Opinion dynamics}

@@

\section{Statistical and social physics}

@@

\section{This thesis: hypotheses and goals}

As said before, this thesis will try to investigate the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we want to determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we would like to investigate the effect of stubborn agents in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms must also be included.\\ 
\newline
In one of their previous papers (\textit{"Modelling opinion dynamics in the age of algorithmic personalization"}, see reference \cite{Perra2019}) prof. Nicola Perra and prof. Luis E.C. Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions. They also considered the effect of nudging (meaning that one opinion is pushed to all agents in the network).\\ Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering coefficient) could result in the formation of echo chambers \cite{Perra2019}. On the other hand, a more heterogeneous contact pattern could hamper this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (namely semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging.\\
They, however, did not investigate the effect of stubborn agents in the network, nor did they investigate networks with a community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have a community structure and relatively low clustering to networks without community structure, but with a high clustering coefficient (a detailed description of concepts like clustering coefficient and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with a high clustering coefficient and, if so, if one can, for example, observe convergence to one opinion inside the communities. We will also pay some attention to networks that have both a high clustering coefficient and a community structure. Can we observe echo chambers in those kinds of networks? If yes, can we determine which property might be the main driver for the formation of echo chambers: high clustering or community structure? \\
% ALSO mention real-world network??
\newline
On the other hand, we want to determine the impact of stubborn agents on the formation and evolution of opinions in the network. Some interesting questions that may come to mind are: Is there an interplay between network structure and stubbornness? What about the possible interplay between the filtering algorithms and the introduction of stubborn people? Does the distribution of stubborn people play a role (eg. randomly distributed versus clustered together inside communities)? \\

@@

The construction of the model is very similar as for the regular stochastic block model. For a SBM-WS with $n_c$ communities of specified sizes, we first construct $n_c$ Watts-Strogatz networks (for each constituent Watts-Strogatz network the rewire probability $\beta$ and the mean degree $K$ are specified), then edges are (with some predefined probability) randomly added between the different Watts-Strogatz communities. The result is a network that can be tuned to have both a community structure and a high clustering coefficient. This is different from the regular stochastic block model that can only account for a community structure and lacks a high clustering.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
Section \ref{networkmodel} gives an overview of the network models that are used in this thesis. In chapter \ref{chap2}, Section \ref{netModel} we already discussed their main properties; here we will mainly discuss why we use these networks. Which of their properties are of interest in this thesis? What are there benefits and possible shortcomings regarding to the goals of this thesis?\\
In section \ref{realnetwork}, we will discuss the real-world networks used in this thesis.\\ % + explain where they come from etc?
Finally, Section \ref{opinionmodel} discusses the opinion dynamics model used in this thesis. Again we will give some focus on possible advantages and disadvantages of the model.

\section{Network models}\label{networkmodel}
\section{Real-world networks}\label{realnetwork}
\section{Opinion dynamics models}\label{opinionmodel}
% use pseudo code here??

>>>>

\selectlanguage{english}
\begin{abstract}
\addcontentsline{toc}{chapter}{Abstract}
The past decade, the rise of social media has become undeniable. More and more people are getting engaged on some type of social media platform and hence, it is probably not wrong to say that nowadays most of us have, beside our real life, a virtual life on social media. This rapidly increased engagement on social media undoubtedly has many advantages. However, every upside has a downside and social media is no different. This thesis deals with the impact of social media on the formation and evolution of opinions in society.\\ 
Social media allows for the spread of information at a faster pace and at an unprecedented scale. However, people are still bounded by time- and cognitive constraints. In order to handle the information overflow, social media companies design filtering algorithms that can fine-tune and individualize the information someone is exposed to. This filtering and individualizing of information may have some potentially dangerous downfalls in the sense that people might get trapped in "echo chambers" that confirm their own beliefs. It is important to know the effects of such algorithmic personalization on the appearance and formation of phenomena such as polarization, spread of extremism, formation of echo chambers, etc.\\
People are, however, more than sheep that thoughtlessly follow their shepherd (contrary to what some books might lead us to believe). They are capable of critical thinking and may be hesitant to change their current opinion. This aspect of potential stubbornness should not be ignored.\\
In this thesis toy models are designed to investigate the interplay between network structure, filtering algorithms and individual resistance to change in the evolution of two competing opinions. We will investigate whether network properties such as community structure and clustering coefficient may enhance or hamper the formation of echo chambers. We will use some basic filtering algorithms to determine their possible effects on the appearance of echo chambers and we will investigate the impact of individual stubbornness. Beside that, we will also pay some attention to the possible ways of distributing the opinions; we will compare randomly distributed cases to cases where entire communities are given one single opinion.
% definetely not finished!! --> add some type of conclusions or not? talk about running model on real world networks?
\end{abstract}

@@

The past decade social media has become ever more important in our daily lives. Nowadays most of us have, beside a real-world life, a virtual life on social media. This increased engagement on social media may have a huge impact on the formation of opinions, not only on the individual scale, but also on a broader, societal scale. \\
It is known that individuals not only form their opinions through self-reflection, but also through interactions with other people and with their surroundings \cite{Perra2019}. The broad range of interactions people undergo on social media with people from all over the world can thus not be underestimated in the process of opinion formation. The understanding of the role of social media on the emergence of polarization and extremism in society is of uttermost importance.\\
\newline
It is also known that some people change their opinions more easily than others, that is, once in a while you encounter somebody that is really stubborn and persistent towards its own opinion. One of the aims of this thesis is to design toy models of opinion dynamics with stubborn actors on theoretical and real-world social networks to analyze the interplay between individual resistance to change and the network structure in the evolution of two competing opinions.\\
\newline
One important feature of opinion dynamics on social media is the use of algorithmic personalization. Social media allow for simultaneous or asynchronous interactions between people without geographical constraints and thus allow for the spread of information at a faster pace and at an unprecedented scale \cite{Perra2019}. People are, however, still bound by temporal and cognitive constraints. In order to assure a more pleasant/convenient experience on their platforms, social media companies use algorithms to order/filter the posts on an individual's time line according to what might be relevant to that individual \cite{Perra2019}. It is of great importance to know the possible effects of these filtering algorithms on the evolution of opinions in the population.\\  
\newline
Opinion dynamics models have two important layers. On one hand, we introduce social networks to describe the underlying structure of social interactions; on the other hand, we need appropriate opinion dynamics/formation models to reproduce the opinion dynamics and opinion formation processes found in real life.\\ Section \ref{complNet} gives an introduction to the first layer, namely that of complex networks. The second layer, the opinion dynamics layer, is introduced in Section \ref{Opinion}. Section \ref{statPhys} introduces concepts such as statistical and social physics. Finally, in Section \ref{goal}, the hypotheses and goals of this thesis are discussed.

\section{Complex networks}\label{complNet}

@@

\section{Opinion dynamics}\label{Opinion}

@@

\section{Statistical and social physics}\label{statPhys}

@@

\section{This thesis: hypotheses and goals}\label{goal}

As said before, this thesis will try to investigate the interplay between the social network structure and the individual resistance to change in the evolution of two competing opinions. On one hand, we want to determine the influence of different network structures on the formation and evolution of opinions, whereas, on the other hand, we would like to investigate the effect of stubborn agents in the network. Since this thesis deals with opinion dynamics on social media, the effect of filtering algorithms must also be included.\\ 
\newline
In one of their previous papers (\textit{"Modelling opinion dynamics in the age of algorithmic personalization"}, see reference \cite{Perra2019}) prof. Nicola Perra and prof. Luis E.C. Rocha investigated both the effect of different filtering algorithms and different network topologies on the formation and evolution of two competing opinions. They also considered the effect of nudging (meaning that one opinion is pushed to all agents in the network).\\ Their main findings were that the algorithmic filtering could not break the status quo when the prevalence of opinions was equally distributed in the population and that topological correlations (such as high clustering coefficient) could result in the formation of echo chambers \cite{Perra2019}. On the other hand, a more heterogeneous contact pattern could hamper this formation of echo chambers \cite{Perra2019}. If the two opinions were, initially, not equally distributed, one of the filtering algorithms (namely semantic filtering) caused a further increase in the predominant opinion \cite{Perra2019}. In the case of nudging, they found that the opinions in the population moved to the nudged opinion relatively fast, even in the case of small nudging \cite{Perra2019}.\\
They, however, did not investigate the effect of stubborn agents in the network, nor did they investigate networks with a community structure. This thesis will build on their model and incorporate these new features. We will compare networks that have a community structure and relatively low clustering to networks without community structure, but with a high clustering coefficient (a detailed description of concepts like clustering coefficient and community structure will be given in Chapter \ref{chap2}, Section \ref{netDef}). We want to investigate whether networks with community structure are able to form echo chambers, such as one can observe in networks with a high clustering coefficient and, if so, if one can, for example, observe convergence to one opinion inside the communities. We will also pay some attention to networks that have both a high clustering coefficient and a community structure. Can we observe echo chambers in those kinds of networks? If yes, can we determine which property might be the main driver for the formation of echo chambers: high clustering or community structure? \\
% ALSO mention real-world network??
\newline
On the other hand, we want to determine the impact of stubborn agents on the formation and evolution of opinions in the network. Some interesting questions that may come to mind are: Is there an interplay between network structure and stubbornness? What about the possible interplay between the filtering algorithms and the introduction of stubborn people? Does the distribution of stubborn people play a role (eg. randomly distributed versus clustered together inside communities)? \\

@@

The construction of the model is very similar to that of the regular stochastic block model. For a SBM-WS with $n_c$ communities of specified sizes, we first construct $n_c$ Watts-Strogatz networks (for each constituent Watts-Strogatz network the rewire probability $\beta$ and the mean degree $K$ are specified), then edges are (with some predefined probability) randomly added between the different Watts-Strogatz communities. The result is a network that can be tuned to have both a community structure and a high clustering coefficient. This is different from the regular stochastic block model that can only account for a community structure and lacks a high clustering.

@@

The previous two chapters, Chapter \ref{chap2} and \ref{chap3}, gave a broad introduction/overview of networks and opinion dynamics models respectively. In this chapter we will summarize the specific networks and opinion dynamics models that are used in this thesis. \\
Section \ref{networkmodel} gives an overview of the network models that are used in this thesis. In chapter \ref{chap2}, Section \ref{netModel} we already discussed their main properties; here we will mainly discuss why we use these networks. Which of their properties are of interest in this thesis? What are there benefits and possible shortcomings regarding to the goals of this thesis?\\
In section \ref{realnetwork}, we will discuss the real-world networks used in this thesis.\\ % + explain where they come from etc?
Finally, Section \ref{opinionmodel} discusses the opinion dynamics model that is implemented in this thesis. Again we will give some focus on possible advantages and disadvantages of the model.

\section{Network models}\label{networkmodel}

This thesis uses three main network models, namely
\begin{itemize}
	\item The Erd\H{o}s-R\'{e}nyi model
	\item The Watts-Strogatz model
	\item The stochastic block model, of which two types are implemented
		\begin{itemize}
			\item[\tiny$\bullet$] The regular stochastic block model (SBM)
			\item[\tiny$\bullet$] The stochastic Watts-Strogatz block model (SBM-WS)
		\end{itemize}
\end{itemize}
Each of these networks serve specific goals which will shortly be discussed in the following paragraphs.\\
\newline
\textbf{The Erd\H{o}s-R\'{e}nyi model}\\
\newline
The Erd\H{o}s-R\'{e}nyi model is mainly used as a benchmark model. In this thesis we focus intensively on two network features that may impact the formation of echo chambers, namely a high clustering coefficient and a community structure. It is therefore necessary to have a model that lacks both of these features to compare the results with. The Erd\H{o}s-R\'{e}nyi model is one of the most simple random network models that doesn't display both of these features and is hence the perfect comparison model. \\
It is expected that the random nature of the model and the lack of a high clustering and a community structure will result in the absence of echo chambers.\\
The Erd\H{o}s-R\'{e}nyi model is also used in the regular stochastic block model to construct the constituent communities.\\
% talk about expected behaviour?? 
\newline
\textbf{The Watts-Strogatz model}\\
\newline
The Watts-Strogatz model is used in this thesis as a general model to investigate the effect of high clustering on the evolution of opinions and the formation of echo chambers.\\
In the paper \textit{"Modelling opinion dynamics in the age of algorithmic personalization"} Prof. Perra and Prof. Rocha already showed that a high clustering coefficient enhances the formation of echo chambers \cite{Perra2019}. Here, we will use the model to investigate the interplay between high clustering and the presence of stubborn people and, on the other hand, we will use it as a comparison for networks that posses a community structure without having a high clustering coefficient. We will investigate whether a community structure behaves similar to a high clustering regarding the appearance of echo chambers.\\
Beside the above, the Watts-Strogatz model is also used to build the constituent communities in the stochastic Watts-Strogatz block model.\\
\newline
\textbf{The stochastic block model}\\
\newline
One of the main features of the stochastic block model is the ability to generate networks with a community structure. We will therefore use this model to investigate the impact of community structure/modularity on the evolution of opinions and the formation of echo chambers.\\
It is expected that a community structure will enhance the formation of echo chambers compared to the random network case. The question remains, however, whether the influence of a community structure on the formation of echo chambers is of the same caliber as that of a high clustering.\\
The stochastic Watts-Strogatz block model is used as a model that has both a high clustering and a community structure. We will check if the combination of both features displays any difference in the formation of echo chambers compared to having only a single feature (either community structure or high clustering).\\
\newline
All the above models have a homogeneous degree distribution. In this thesis, we didn't focus on network models which generate networks with a heterogeneous degree distribution such as the Barab\'{a}si-Albert model. It might, however, be an interesting extension to investigate such types of models as well.
% NOT FINISHED: talk about shortcommings --> models, neither of them describes real networks completely accurate, etc

\section{Real-world networks}\label{realnetwork}
\section{Opinion dynamics models}\label{opinionmodel}
% use pseudo code here??
